---
title: "PheWAS Associations with Hydronephrosis"
author: Siwei Zhang
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
  style_switcher: false
  thumbnails: no
  lightbox: yes
  gallery: no
  highlight: tango
  use_bookdown: no
  toc_depth: 6
  fig_caption: yes
  embed_fonts: no
  keep_md: no
  number_sections: yes
# bibliography: references.bib
# csl: apa-single-spaced.csl
---

```{r, include = FALSE}
# knitr::opts_chunk$set(
#   collapse = TRUE,
#   comment = "#>"
# )
embed_png <- function(image_title, width = "100%") {
  knitr::asis_output(
    glue::glue("<div style=\"text-align: center;\">",
                 "<img src=\"{image_title}\" width={width} />",
               "</div>")
  )
}
```

# Introduction

- PheMIME$^1$ enables users to explore phenome-wide causal associations. We utilized hydronephrosis as a case study to exemplify the effectiveness of the PheMIME in uncovering the established causes of hydronephrosis while also suggesting less common causes of hydronephrosis.


# PheMIME analysis

The discovery process of phenome-wide causal associations in hydronephrosis was conducted through the 'Reproducible Multimorbidities Exploration' and 'Multimorbidity Similarities Exploration' modules in PheMIME.

- **Robust and Consistent Comorbidities**:

  The robust and consistent comorbidities associated with hydronephrosis, which included well-known causes such as **obstructing stones in the ureter**, **vesicoureteral reflux**, and **congenital defects were identified**.

- **Less Common Causes** 

  Additionally, less common causes of hydronephrosis were discovered, including **certain cancers**, **urethral stricture**, and **abnormal renal vasculature**.

- **Complex Causative Patterns**

  Our analysis also revealed complex causative patterns of conditions related to hydronephrosis, such as **pyelonephritis** and **E. coli infection**, consistent with the known knowledge that obstruction and stagnation of urine is presumed to be the reason for predisposition to pyelonephritis and E. coli contributes as a predominant pathogen to the etiology of pyelonephritis and urinary tract infections. However, we also noticed multiple types of conditions related to **mineral metabolism**, likely to have a possible genetic component that could predispose to renal calculi and hydronephrosis. These unexpected findings underscore the potential of PheMIME to reveal novel disease associations for further investigation.

### Reproducible Multimorbidities Exploration

- Identify the most consistent and top-co-occurred clinical disease phenotypes based on comorbidity strength from the Manhattan and Scatter plot through the module of **Reproducible Multimorbidities Exploration**. 

```{r, echo = FALSE}
embed_png("hydronephrosis_multimorbidity.png")
```

### Multimorbidity Similarities Exploration

- Identify the most consistent and top clinical disease phenotypes with significant multimorbidity similarities (how these phenotypes connect to other phecodes in the network) through the module of **Multimorbidity Similarities Exploration**. 

```{r, echo = FALSE}
embed_png("hydronephrosis_similarity.png")
```

# References

1. Zhang, S., Strayer, N., Vessels, T., Choi, K., Wang, G. W., Li, Y., Bejan, C. A., Hsi, R. S., Bick, A. G., Velez Edwards, D. R., Savona, M. R., Philips, E. J., Pulley, J., Self, W. H., Hopkins, W. C., Roden, D. M., Smoller, J. W., Ruderfer, D. M., & Xu, Y. (n.d.). PheMIME: An Interactive Web App and Knowledge Base for Phenome-Wide, Multi-Institutional Multimorbidity Analysis. https://doi.org/10.1101/2023.07.23.23293047












