---
title: ""
author: "Siwei Zhang"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute: null
  style_switcher: no
  thumbnails: no
  lightbox: yes
  gallery: no
  highlight: tango
  use_bookdown: no
  toc_depth: 6
  fig_caption: yes
  embed_fonts: no
  keep_md: no
  number_sections: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
embed_png <- function(image_title, width = "100%") {
  knitr::asis_output(
    glue::glue("<div style=\"text-align: center;\">",
                 "<img src=\"{image_title}\" width={width} />",
               "</div>")
  )
}
```

## Motivation

Phenome-based disease multimorbidity derived from EHR is becoming essentially important in precision medicine. It is useful to prevent multimorbid conditions and provide evidence-based treatment to patients by finding pathogenesis of disease, providing insights for disease prognosis and identifying underlying shared molecular mechanisms. And identifying disease multimorbidity provides a new opportunities of clinical trials design and drug repurposing by recruiting patients with complex disease profiles that more closely resembles the real-world and using single drug to treat multiple conditions concurrently. However, most of the previous studies focus on a single EHR system due to the limitation of sharing individual-level data across different EHR systems and there lacks a common visualization tool to conduct statistical analysis and make inference on latent structure of disease multimorbidity clustering and causation. In this research, we aim to develop a visualization tool that can be used by clinicians and researchers to explore multimorbidity in EHR data from multiple institutions. The tool will be able to visualize the reproducibility of multimorbidity across different EHR systems and identify the latent disease multimorbidity clustering and causation given the condition of statistical significance. 

## Results

We present Phecode-wide Multi-Institutional Multimorbidity Explorer  (PheMIME) which is an interactive visualization tool of phenome-based disease network. By utilizing summary statistics from association analysis to represent the disease multimorbidity strength, a database was collected and computed from association analysis across three institutions: Vanderbilt University Medical Center, Massachusetts General Brigham, and UK Biobank. By framing the association analysis of disease multimorbidity as a network problem, it allows the analysis and inference of the latent disease phenotypes clustering, which extends the pairwise multimorbidity to the multivariate disease clusters. The tool takes an input from the selection of the interested disease phenotype and generates plots interactive with the table that either display the pairwise multimorbidity strength within single system or show the consistency of multimorbidity across systems, which gives users the freedom choosing disease phenotypes based on prior knowledge or doing analysis of adding or removing disease phenotypes according to the discovery by associationSubgraphs. The new version of associationSubgraphs embedded into the tool interactive with plots and tables allows tracking the trajectory of forming disease multimorbidity. Finally, the implementation with a use-case in Schizophrenia is demonstrated.

## Application

The PheMIME is available at https://prod.tbilab.org/PheMIME/.

```{r, echo = FALSE}
embed_png("overall_original.png")
```

## Report Problems

Your feedback is valuable to us! Please share your comments and suggestions by submitting an issue on our GitHub page [Github Issues](https://github.com/tbilab/PheMIME/issues)
