---
title: "Multi-Institutional Disease Multimorbidity Explorer Detailed Overview"
author: Siwei Zhang
date: "`r Sys.Date()`"
output:
    rmdformats::downcute:
    style_switcher: false
    thumbnails: no
    lightbox: yes
    gallery: no
    highlight: tango
    use_bookdown: no
    toc_depth: 6
    fig_caption: yes
    embed_fonts: no
    keep_md: no
    number_sections: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
embed_png <- function(image_title, width = "100%") {
  knitr::asis_output(
    glue::glue("<div style=\"text-align: center;\">",
                 "<img src=\"{image_title}\" width={width} />",
               "</div>")
  )
}
```


## Why this tool is in need?

- Disease multimorbidity

Scientific question is in need: Identifying disease multimorbidity is important in both of preventing multimorbid conditions and providing precision treatment to patients. Furthermore, it will be helpful to re-define the diagnosis system with more precision. Besides, getting more in-depth knowledge of disease multimorbidity makes the opportunity of recruiting patients with more complex multiple disease profiles that more closely resembles the real-world situation into reality and provides a new opportunity of drug repurposing by using single drug to treat multiple conditions concurrently.

- Integration of different population cohort

EHR-based observational studies are usually based on a single institution, while different institutions may have diverse populations and different healthcare settings, and within each institution, it might also be biased that physicians have a different experience or systematic error when recording diagnoses, which may lead to different disease multimorbidity patterns and limits the generalizability of the results. In addition, it is difficult to conduct a large-scale study to identify and reproduce disease multimorbidity across different institutions. Therefore, integrating disease multimorbidity across different institutions is the challenging task due to the heterogeneity of EHR data. 

- Efficient and reliable visualization 

There lacks a common visualization tool to help conduct statistical analysis and make inference of the latent structure of disease multimorbidity clustering and causation. The limitation of sharing individual-level EHR data, technical computing resources, and ethical reasons challenges the research on creating and building a common standard visualization tool to better help the clinical discoveries. An efficient visualization tool based on large-scale database will provide a comprehensive and more valuable resource for validating known phenome-based multimorbidity and discovering unknown multimorbidity, and meanwhile could increase statistical power. Most EHR-based phenome-wide approaches focused on studying each phenotype independently, but our approach explored phenome-wide multimorbidity patterns leveraging large-scale EHR systems and our multimorbidity networks database provide deeper understanding of differences across institutions. 


## How to use this tool?

### Data integration for three institutions

  You do not need to upload data and the internal database is stored in the bottom layer of this tool. 
  
- Three databases including the individual-level data for 250,000 randomly selected patients from VUMC and MGB EHR systems, and 431,105 patients from UKB in-hospital records were gathered. As for VUMC and MGB, each Phecode was denoted as an occurrence if occurred two or more times in the patients’ collapsed record. As for UKB, the ICD 10 codes were mapped to Phecodes and each Phecode was denoted as an occurrence if occurred at least one time in patients’ collapsed record. 

- Logistic regressions adjusted by patient age at extraction date, EHR age (patient age at last recorded visit), sex, race, and EHR burden (number of unique Phecodes present in patients’ records) were run for each pair of two phecodes with two conditions of either phecode A or phecode B regarded as the outcome and the other one treated as the independent variable. The averaged test statistic is used to characterize the comorbidity strength for all pairs. And Pearson correlation of the common comorbidity between two comorbidity patterns is used as the similarity. The summarized data from three institutions was stored into our database. 

### User input and selection

- You could select EHR system to compare the consistency in the below "Institutions". You could also use multimorbidity strength from single institution or combined multimorbidity strength from multiple institutions by selecting which EHR system to use. 

- You could input statistical significance level in the below box such as "Input p-value"

- Whenever you select a new institution, or you input another significance level, you need click "Update" button in each page. It is a two-step analysis that will avoid transient changes when changing the input values.

```{r, echo = FALSE}
embed_png("consistency_selection.png")
```
  
**Explore Multimorbidity**

- You could hover, click, double-click or brush the points in the plots. When you hover the point, the information of the points will be shown. Brushing a region to include the points you want to select and you can add new points by clicking the point or remove the points by double-click the point.

- Table is interactive with the plots. You can check a phenotype's consistency across different systems or multimorbidity strength by clicking the row corresponding to this phenotype in the table and then clicking "Update" button to update the plots. 

- Whenever there is a change in the plots or in the table, click "update associationSubgraphs" to update the network visualization

```{r, echo = FALSE}
embed_png("explore_page.png")
```


## Reference

Aguado,A. et al. (2020) MorbiNet: multimorbidity networks in adult general population. Analysis of type 2 diabetes mellitus comorbidity. Sci Rep, 10, 2416.

Bastarache,L. et al. (2022) Phenome-Wide Association Studies. JAMA, 327, 75.

Bojko,A. (2009) Informative or misleading? Heatmaps deconstructed. In, Lecture Notes in Computer Science (including subseries Lecture Notes in Artificial Intelligence and Lecture Notes in Bioinformatics)., pp. 30–39.

Hassaine,A. et al. (2020) Learning multimorbidity patterns from electronic health records using Non-negative Matrix Factorisation. J Biomed Inform, 112, 103606.

Kuan,V. et al. (2023) Identifying and visualising multimorbidity and comorbidity patterns in patients in the English National Health Service: a population-based study. Lancet Digit Health, 5, e16–e27.

Chong,L.W. et al. (2018) Association of viral hepatitis and bipolar disorder: A nationwide population-based study. J Transl Med, 16.





