---
title: "Multi-Institutional Disease Multimorbidity Explorer Usage Example"
author: Siwei Zhang
date: "`r Sys.Date()`"
output:
    rmdformats::downcute:
    style_switcher: false
    thumbnails: no
    lightbox: yes
    gallery: no
    highlight: tango
    use_bookdown: no
    toc_depth: 6
    fig_caption: yes
    embed_fonts: no
    keep_md: no
    number_sections: yes
---
<!-- <style type="text/css"> -->
<!-- .main-container { -->
<!-- max-width: 3000px; -->
<!-- margin-left: auto; -->
<!-- margin-right: auto; -->
<!-- } -->
<!-- </style> -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
embed_png <- function(image_title, width = "100%") {
  knitr::asis_output(
    glue::glue("<div style=\"text-align: center;\">",
                 "<img src=\"{image_title}\" width={width} />",
               "</div>")
  )
}
```

# Introduction

Here we describe a sample use-case of the Multi-Institutional Disease Multimorbidity Explorer for investigating the results of the diseasome-wide multimorbidities analysis corresponding to Schizophrenia. We show step-by-step that the utility of the Multi-Institutional Disease Multimorbidity Explorer and how it can conduct analysis in a more efficient way to identify new discoveries of unknown disease associations and make inference of the causal path toward specific diseases, and provide an approach to compare and validate the reproducibility of multimorbidity across multiple institutions. 

# Main app

## Loading app and making selection

When the app is loaded and there is a Phecode table on the first page, it provides the opportunity for users to select their interested phenome to discover. Once they selected a phecode, the following page will load the corresponding analysis results across multiple institutions and they will update according to any new selection of a phecode on the first page so that users can also discover other interested phenotypes one by one. In this case, we select __Schizophrenia__.

```{r, echo = FALSE}
embed_png("selection.png")
```

## Consistency across institutions

- By adjusting p-value threshold, the consistency pattern between any two institutions for all independent pairwise disease phenotypes is shown and the pattern for those phenotypes co-occurred with selected phenotype is presented as well. If the patterns stay similar across three systems, then there is strong evidence support of existing the disease multimorbidity that discovered in the next several pages. 

- Points in the plot represent the multimorbidity strengths between a pair of phecodes that each axis value corresponds to the strength in each of the institution. P-value bar is used to adjust the threshold that points corresponding to any pair of phecodes with p-value smaller than the threshold would be included in the plot. Whenever the selection of institutions or the p-value threshold are updated, or the user selects another phenotype to discover, clicking the "update consistency plot" to update the plot.

```{r, echo = FALSE}
embed_png("consistency_selection.png")
```

- Blue points denote multimorbidity scores corresponding to all pairs of phecodes and green points only show phecode pairs including the selected phecode with p-value smaller than the threshold. We can see for __Schizophrenia__, overall the consistency between VUMC and MGH performs better than the consistency between VUMC and UK Biobank, which indicates the disease multimorbidity discovery might be more trustworthy using combined VUMC and MGH database.

```{r, echo = FALSE}
embed_png("consistency_vumc_mgh.png")
```

```{r, echo = FALSE}
embed_png("consistency_vumc_ukbb.png")
```

## Overall multimorbidity network

Users can visualize the high-dimensional summary statistics from the association analysis in a more efficient way to identify new discoveries of unknown disease associations. In addition, the conditional independence that the association of phenome a and phenome b is due to phenome c they both are connected to could be validated as well through network visualization across different systems. We demonstrate the overall multimorbidity discovery using __Schizophrenia__ example. 

After Bonferroni correction for 1,701 tests (p < $2.94*10^{-5}$) and given the multimorbidity strength threshold at 43.47, it is not surprised that there are a lot of phecodes in the mental disorders category including psychosis, suicidal ideation, bipolar disorder, suicidal ideation or attempt, conduct disorders, mood disorders, and hallucinations significantly associated with __Schizophrenia__. And another significant and non-psychiatric group is __neurological__ including eepilepsy, recurrent seizures, convulsions and so on. The second group of non-psychiatric multimorbidities was __injuries and poisonings__: poisoning by psychotropic agents; poisoning by anticonvulsants and anti-Parkinsonism drugs; adverse drug events and drug allergies; and toxic effect of other substances, chiefly nonmedicinal as to source. 

By viewing the phenotypes that significantly co-occurred with __Schizophrenia__ which denoted with colors in the network plot, it is possible to infer the causal path of some disease phenotypes. For example, if looking at the data from VUMC, after setting significance level at $2.94*10^{-5}$ and strength threshold at $39.661$, we can see infectious disease of __viral hepatitis__ and __viral hepatitis C__ are significantly associated with __Schizophrenia__ and are clustered into the same group, while the relationship is that __viral hepatitis__ does not directly cause __Schizophrenia__ but they are connected to __Schizophrenia__ through __Biporlar__. One of the possible reasons is that the adverse influence of __viral hepatitis__ cause the brain to be more vulnerable to neuroinflammation and therefore increases the risk of __Biporlar__, which lead to __Schizophrenia__. We can see the similar causal path when using MGB data when setting significance level at $2.94*10^{-5}$ and strength threshold at $38.126$.

```{r, echo = FALSE}
embed_png("multimorbidity_network_overall_sch.png")
```

```{r, echo = FALSE}
embed_png("multimorbidity_overall_inf_mgb.png")
# embed_png("multimorbidity_overall_psy.png")
```

```{r, echo = FALSE}
embed_png("multimorbidity_overall_inf_vumc.png")
# embed_png("multimorbidity_overall_nonpsy.png")
```

We can also directly see the improvement of utilizing this tool compared with the static manhattan plot. The static manhattan plot is hard to visualize all significant phenotypes associated with __Schizophrenia__. And it does not contain the hidden information of disease subgraphs. It raises a question as well, what if the users only want to select some of their interested phenotypes among all these phenotypes? The next page of "Explore multimorbidity" provides a more detailed application on it.

```{r,echo=FALSE}
embed_png("sch_manhattan.png")
```

## Explore multimorbidity

- Both the interactive version of Manhattan and Scatter plot are provided. The selected phenotypes in Manhattan plot will also be chosen automatically in Scatter plot such that the magnitude and consistency of the disease multimorbidity or similarity could be visualized at the same time. If the disease multimorbidity with large magnitude also presents high consistency across different systems, then there is strong evidence of existing such disease multimorbidity.

- The associationSubgraph incorporated with user-selected phenomes colored provides an interactive visualization method to quickly explore and make inference of the latent structure of disease clusters. By annotating the disease phenotypes in the static Manhattan/Scatter plot, the disease category and corresponding multimorbidity strengths in three institutions are highlighted in the table. Furthermore, a modified version of the associationSubgraph helps track the trajectory of forming the disease subgroups. The nodes which represent phenotypes in the network will be categorized into two groups with annotated phenomes colored by their disease categories and the other nodes without annotation colored into grey which makes the analysis and inference of disease multimorbidity into reality. By selecting the phenomes highly co-occurred with the interested disease and validating such phenomes that consistent across systems, the selected phenotypes might be clustered into separate subgroups, which represents different mechanisms. 

```{r,echo=FALSE}
embed_png("explore_page.png")
```

### Schizophrenia

- By selecting the most associated and highly consistent phenotypes in the plots across two institutions of VUMC and MGB, and such phenotypes were denoted with colors in the network plot. We can see the similar causal relationship path of __Schizophrenia__. We can see infectious disease of c and __viral hepatitis C__ are highly associated with __Schizophrenia__ and are clustered into the same group, while the relationship is that __viral hepatitis__ does not directly cause __Schizophrenia__ but they are connected to __Schizophrenia__ through __Substance addiction and disorders__ and __Biporlar__. One of the possible reasons is that the adverse influence of __viral hepatitis__ cause the brain to be more vulnerable to neuroinflammation and therefore increases the risk of __Biporlar__, which lead to __Schizophrenia__.

```{r,echo=FALSE}
embed_png("explore_plot_inf.png")
```

```{r,echo=FALSE}
embed_png("explore_network_inf.png")
```

- And we can also see that some disease phenotypes in __digestive__ category that not selected in the plots but grouped in the same cluster with __viral hepatitis__, __Biporlar__ and __Schizophrenia__. We can search the __digestive__ phenotypes in the table and click on the rows corresponding to these phenotypes then click "update manhattan/scatter plot".

```{r,echo=FALSE}
embed_png("digestive_table.png")
```

- Finally, click "update associationsubgraphs" and we can see the __digestive__ phenotypes are also highlighted in the network plot

```{r,echo=FALSE}
embed_png("digestive_network.png")
```

## Reference

Aguado,A. et al. (2020) MorbiNet: multimorbidity networks in adult general population. Analysis of type 2 diabetes mellitus comorbidity. Sci Rep, 10, 2416.

Bastarache,L. et al. (2022) Phenome-Wide Association Studies. JAMA, 327, 75.

Bojko,A. (2009) Informative or misleading? Heatmaps deconstructed. In, Lecture Notes in Computer Science (including subseries Lecture Notes in Artificial Intelligence and Lecture Notes in Bioinformatics)., pp. 30–39.

Hassaine,A. et al. (2020) Learning multimorbidity patterns from electronic health records using Non-negative Matrix Factorisation. J Biomed Inform, 112, 103606.

Kuan,V. et al. (2023) Identifying and visualising multimorbidity and comorbidity patterns in patients in the English National Health Service: a population-based study. Lancet Digit Health, 5, e16–e27.

Chong,L.W. et al. (2018) Association of viral hepatitis and bipolar disorder: A nationwide population-based study. J Transl Med, 16.



