---
title: "Multi-Institutional Multimorbidity Explorer Usage Example"
author: Siwei Zhang
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
  style_switcher: false
  thumbnails: no
  lightbox: yes
  gallery: no
  highlight: tango
  use_bookdown: no
  toc_depth: 6
  fig_caption: yes
  embed_fonts: no
  keep_md: no
  number_sections: yes
---
<!-- <style type="text/css"> -->
<!-- .main-container { -->
<!-- max-width: 3000px; -->
<!-- margin-left: auto; -->
<!-- margin-right: auto; -->
<!-- } -->
<!-- </style> -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
embed_png <- function(image_title, width = "100%") {
  knitr::asis_output(
    glue::glue("<div style=\"text-align: center;\">",
                 "<img src=\"{image_title}\" width={width} />",
               "</div>")
  )
}
```

# Introduction

Here we describe a sample use-case of the Multi-Institutional Disease Multimorbidity Explorer for investigating the results of the diseasome-wide multimorbidities analysis corresponding to Schizophrenia. We show step-by-step that the utility of the Multi-Institutional Disease Multimorbidity Explorer and how it can conduct analysis in a more efficient way to identify new discoveries of unknown disease associations and make inference of the causal path toward specific diseases, and provide an approach to compare and validate the reproducibility of multimorbidity across multiple institutions. 

# Main app

### Disease Selection

When the app is loaded and there is a Phecode table on the first page, it provides the opportunity for users to select their interested phenome to discover. Once they selected a phecode, the following page will load the corresponding analysis results across multiple institutions and they will update according to any new selection of a phecode on the first page so that users can also discover other interested phenotypes one by one. In this case, we select __Schizophrenia__.

```{r, echo = FALSE}
embed_png("selection.png")
```

### Multimorbidity Consistency Inspection

- By adjusting p-value threshold, the consistency pattern between any two institutions for all independent pairwise disease phenotypes is shown and the pattern for those phenotypes co-occurred with selected phenotype is presented as well. If the patterns stay similar across three systems, then there is strong evidence support of existing the disease multimorbidity that discovered in the next several pages. 

- Points in the plot represent the multimorbidity strengths between a pair of phecodes that each axis value corresponds to the strength in each of the institution. P-value bar is used to adjust the threshold that points corresponding to any pair of phecodes with p-value smaller than the threshold would be included in the plot. Whenever the selection of institutions or the p-value threshold are updated, or the user selects another phenotype to discover, clicking the "update consistency plot" to update the plot.

```{r, echo = FALSE}
embed_png("consistency_selection.png")
```

- Blue points denote multimorbidity scores corresponding to all pairs of phecodes and green points only show phecode pairs including the selected phecode with p-value smaller than the threshold. We can see for __Schizophrenia__, overall the consistency between VUMC and MGH performs better than the consistency between VUMC and UK Biobank, which indicates the disease multimorbidity discovery might be more trustworthy using combined VUMC and MGH database.

```{r, echo = FALSE}
embed_png("consistency_vumc_mgh.png")
```

```{r, echo = FALSE}
embed_png("consistency_vumc_ukbb.png")
```


### Multimorbidity Network Visualization

Users can visualize the high-dimensional summary statistics from the association analysis in a more efficient way to identify new discoveries of unknown disease associations. By integrating a dynamic network visualization and clustering methodology called associationSubgraphs, users can examine and compare the multimorbidity networks constructed based on the multimorbidity strength measurements. This module permits exploration of the network's subgraph structures and dynamic clustering for any multimorbidity network from a single institution or an amalgamation of multiple institutions. Moreover, this module enables users to apply filters and emphasize any significant multimorbidities and investigate their interconnections and enriched subgraphs. We demonstrate the discovery using __Schizophrenia__ example. 

After Bonferroni correction for 1,701 tests (p < $2.94*10^{-5}$) and given the multimorbidity strength threshold at 43.47, it is not surprised that there are a lot of phecodes in the mental disorders category including psychosis, suicidal ideation, bipolar disorder, suicidal ideation or attempt, conduct disorders, mood disorders, and hallucinations significantly associated with __Schizophrenia__. And another significant and non-psychiatric group is __neurological__ including eepilepsy, recurrent seizures, convulsions and so on. The second group of non-psychiatric multimorbidities was __injuries and poisonings__: poisoning by psychotropic agents; poisoning by anticonvulsants and anti-Parkinsonism drugs; adverse drug events and drug allergies; and toxic effect of other substances, chiefly nonmedicinal as to source. 

By viewing the phenotypes that significantly co-occurred with __Schizophrenia__ which denoted with colors in the network plot, we discovered a strong comorbidity with Viral Hepatitis. For example, if looking at the data from VUMC, after setting significance level at $2.94*10^{-5}$ and strength threshold at $39.661$, we can see infectious disease of __viral hepatitis__ and __viral hepatitis C__ are significantly associated with __Schizophrenia__ and are clustered into the same group, while the causal relationship between __viral hepatitis__ and __Schizophrenia__ is unknown but they are connected to __Schizophrenia__ through __Biporlar__. One of the possible reasons is that the adverse influence of __viral hepatitis__ cause the brain to be more vulnerable to neuroinflammation and therefore increases the risk of __Biporlar__, which lead to __Schizophrenia__. We can see the similar path when using MGB data when setting significance level at $2.94*10^{-5}$ and strength threshold at $38.126$.

```{r, echo = FALSE}
embed_png("multimorbidity_network_overall_sch.png")
```

#### VUMC

```{r, echo = FALSE}
embed_png("multimorbidity_overall_inf_mgb.png")
# embed_png("multimorbidity_overall_psy.png")
```

#### MGB

```{r, echo = FALSE}
embed_png("multimorbidity_overall_inf_vumc.png")
# embed_png("multimorbidity_overall_nonpsy.png")
```

### Improvement compared to static Manhattan plot

We can also directly see the improvement of utilizing this tool compared with the static manhattan plot. The static manhattan plot is hard to visualize all significant phenotypes associated with Schizophrenia. And it does not contain the hidden information of disease subgraphs. It raises a question as well, what if the users only want to select some of their interested phenotypes among all these phenotypes? The next page of “Explore multimorbidity” provides a more detailed application on it.

```{r,echo=FALSE}
embed_png("sch_manhattan.png")
```


## Explore multimorbidity

PheMIME allows users to use phenome-wide multimorbidity patterns of __Schizophrenia__ for exploration and network analysis. 

- PheMIME generates an interactive Manhattan plot, enabling users to select phecodes based on their pairwise multimorbidity strength. A scatter plot for comparing between two systems is also generated, helping users to select robust multimorbidities that exhibit reproducible patterns across systems. The interactive features of the Manhattan and Scatter plots enable the selection and highlighting of a consistent set of the same phecodes, based on the magnitude (Manhattan plot) and consistency (Scatter plot) of disease multimorbidities. For instance, the Manhattan plot might reveal phecodes highly co-occurring with __Schizophrenia__, while the Scatter plot might illustrate phecodes highly consistent across institutions. 

- The data table shows comorbid phecodes of __Schizophrenia__, its description, disease categories and corresponding multimorbidity strengths among three institutions. This table is interactive with the Manhattan and Scatter plots, allowing users to add or remove phecodes by clicking on the rows in the table. If a disease multimorbidity exhibits both a large magnitude and high consistency across different systems, it strongly indicates a robust disease multimorbidity across the systems. 

- The enhanced version of associationSubgraphs provides an interactive visualization to quickly explore subgraph structure of __Schizophrenia__ multimorbidities in the selected phecodes that are both significant and reproducible across institutions. Network nodes are annotated into two groups, with the selected phecodes color-filled based on disease categories and the other unselected phecodes (nodes) color-filled in grey. As expected, a major subgraph enriching the selected __Schizophrenia__ multimorbidity phecodes is mental disorders, with other subgraphs showing enrichment in infectious diseases and neoplasms.

```{r,echo=FALSE}
embed_png("overall.png")
```

- Users can also conduct comparative analyses of multimorbidity patterns across different populations. By exploring the interactive Manhattan and scatter plot compared Vanderbilt University Medical Center (VUMC) with Massachusetts General Brigham (MGB), VUMC with UK Biobank (UKB) and MGB with UKB, we discovered a strong comorbidity of __Schizophrenia__ with __Viral Hepatitis B__ and __Viral Hepatitis C__. From the below figures, __Viral Hepatitis__, __Viral Hepatitis B__ and __Viral Hepatitis C__ multimorbidity exhibit both large magnitudes and high consistencies across different systems, it strongly indicates a robust disease multimorbidity across the systems. 

#### VUMC vs MGB

```{r,echo=FALSE}
embed_png("hepatitis_vumc_mgb.png")
```

#### VUMC vs UKB

```{r,echo=FALSE}
embed_png("hepatitis_vumc_ukb.png")
```

#### MGB vs UKB

```{r,echo=FALSE}
embed_png("hepatitis_mgb_ukb.png")
```

- Users can further examine in detail of multimorbidity patterns and it is discovered that there is more consistency of comorbidity within the patient cohorts of VUMC and MGB, with the comorbidity intensities being notably stronger in these cohorts as compared to the general UK population represented by the UK Biobank (UKB) cohort, especially in the case of __Viral Hepatitis B__. This distinction could underscore the heterogeneity effect due to population differences and, as the exact reasons and mechanisms behind __Viral Hepatitis__ and __Schizophrenia__ remain elusive, the observed disparities among different cohorts might provide intriguing insights, pave the way for additional investigation, potentially leading to new understanding of the mechanisms and elucidating the impact of population differences on disease multimorbidity patterns.

```{r,echo=FALSE}
embed_png("hepatitis_table.png")
```






